---
title: "Describing Continuous Data: Center, Spread, Shape, and Interval Probability"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"

  
format: pptx

execute:
  echo: true
  warning: false
  message: false
---


```{r, echo=FALSE}
library(tidyverse)
library(scales)
```

### Workflow for Describing Continuous Data

::: {style="font-size:0.68em"}
-   describing continuous data before modeling involves a simialr sequence of steps as described for discrete values:

    1.  define the observational unit\
    2.  identify the variable as continuous\
    3.  determine summaries of center and spread\
    4.  evaluate distribution shape across the measurement scale\
    5.  select visualizations that preserve numerical structure and distribution form

-   numerical summaries and graphical displays with how quantitative values are recorded and help describe what the measurements represent across patients
:::

------------------------------------------------------------------------

### Workflow and Descriptive Interpretation

::: {style="font-size:0.68em"}
Each step in the workflow defines a component of continuous measurements described by statistical summaries or plots:

-   `observational unit`: determines what one recorded value represents within the study design\
-   `continuous variable`: defines the measurement as a quantitative value on a numerical scale\
-   `summary`: quantifies central tendency and dispersion\
-   `distribution analysis`: evaluates symmetry, skewness, tails, and concentration across the scale\
-   `probability representation`: expresses how much of the sample falls within domain-specific ranges\
-   `visualization`: encodes the distribution in a form that preserves scale, shape, and comparability
:::

------------------------------------------------------------------------

### Define the Observational Unit

::: {style="font-size:0.67em"}
Review:

-   observational unit is the entity represented by one row of data
-   each observational unit contributes one numerical value measured on a quantitative scale

Continuous examples of observational units include:

-   one patient encounter\
-   one blood draw\
-   one biopsy specimen\
-   one laboratory assay result\
-   one follow-up visit
:::

------------------------------------------------------------------------

### Identify Continuous Variables

::: {style="font-size:0.67em"}
After the observational unit is established, continuous variables are recognized by how values are recorded:

-   values are quantitative rather than labels\
-   values can vary across a numerical scale\
-   values may take many distinct magnitudes within an interval\

Examples in biomedical data include:

-   cytokine concentration\
-   blood pressure\
-   serum creatinine\
-   tumor size

Descriptive summaries for continuous data help determine:

-   where values are centered\
-   how widely values vary\
-   how values are distributed across the scale
:::

------------------------------------------------------------------------

### Determine the Data Summary for Continuous Variables

::: {style="font-size:0.67em"}
-   continuous variables are commonly described with summaries of center, spread, and shape :

    -   `mean` : arithmetic center of the sample\
    -   `median` : middle ordered value\
    -   `standard deviation` : variability around the sample mean\
    -   `interquartile range` : spread of the middle 50% of observations\
    -   `minimum and maximum` : observed range endpoints\
    -   `quantiles` : positional summaries along the ordered scale

-   no single summary fully describes a continuous distribution

-   valid description combines numerical summaries with plots that show how values occupy the measurement scale
:::

------------------------------------------------------------------------

### Continuous Distributions Describe Measurement Structure

::: {style="font-size:0.67em"}
-   continuous distributions describe how measured values are organized across the numerical scale

-   descriptive statistical analysis involves summarizing how values are distributed across observational units without removing the scale on which the measurements were recorded

For continuous variables, descriptive summaries show:

-   **center** → where values tend to cluster along the scale

-   **spread** → how widely values vary around the center

-   **shape** → whether the distribution is symmetric, skewed, heavy-tailed, or multi-peaked

-   **range and tails** → whether observations extend into unusually low or high values
:::

------------------------------------------------------------------------

### Probability Quantifies Where Continuous Values Fall

::: {style="font-size:0.67em"}
For continuous data, descriptive probability is often expressed over an interval of values rather than at one exact value

For a sample of size $n$:

$$
\hat{P}(a \le X \le b) = \frac{\text{number of observations with values between } a \text{ and } b}{n}
$$

Formula defines the probability that a measured value falls within the interval from $a$ to $b$

Where:

-   $X$ : continuous variable\
-   $a$ : lower bound of the interval\
-   $b$ : upper bound of the interval\
-   $n$ : total number of observations
:::

------------------------------------------------------------------------

### Visualizations for Continuous Data

::: {style="font-size:0.67em"}
-   plots are chosen to preserve the numerical scale and reveal how values occupy that scale

Visualizations for continuous variables include:

-   `histogram` : shows how observations accumulate across intervals of the scale\
-   `density plot` : shows a smooth estimate of distribution shape\
-   `boxplot` : summarizes median, quartiles, and tail extension\
:::

------------------------------------------------------------------------

### Data Overview: Continuous Biomarkers in Respiratory Infection

::: {style="font-size:0.67em"}
**Background**: Suppose we observe inflammatory biomarkers collected from a cohort of patients evaluated for respiratory infection

Each row represents one patient visit, and each variable records a continuous biological measurement:

-   `il6` : interleukin-6 concentration in \[pg/mL\]\
-   `tnf_alpha` : tumor necrosis factor-alpha concentration in \[pg/mL\]\
-   `neutrophils` : neutrophil count in units of \[$10^9$/L\]

Research Question:

> How are these biomarker values distributed across patients, and what do center, spread, and shape indicate about the observed sample?
:::

------------------------------------------------------------------------

### Simulated Biomarker Dataset

::: {style="font-size:0.64em"}
-   dataset of 220 patient-level observations, where each row corresponds to a single clinical visit and each column represents a continuous biomarker measurement

```{r}
set.seed(303)

biomarker_df <- tibble(
  patient_id = sprintf("B%03d", 1:220),
  il6 = round(rlnorm(220, meanlog = log(12), sdlog = 0.70), 2),
  tnf_alpha = round(rlnorm(220, meanlog = log(18), sdlog = 0.45), 2),
  neutrophils = round(rnorm(220, mean = 5.9, sd = 1.3), 2)
)

biomarker_df |> slice_head(n = 8)
```
:::

------------------------------------------------------------------------

### Mean and Median: Two Summaries of Center

::: {style="font-size:0.67em"}
**Mean** is expressed as:

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i
$$

Average of all observed values in the sample

Where:

-   $\bar{x}$ : sample mean\
-   $n$ : number of observations\
-   $x_i$ : observed value for the $i$th observational unit\
-   $\sum_{i=1}^{n} x_i$ : total of all observed values

**Median** : middle value after the sample is ordered from smallest to largest

-   mean uses the magnitude of every value\
-   median uses the order of values in the sample
:::

------------------------------------------------------------------------

### Standard Deviation and Interquartile Range

::: {style="font-size:0.67em"}
**Standard deviation**:

-   measure that captures variability around the mean
-   summary of dispersion that quantifies how far observations extend from the mean across the entire distribution

$$
s = \sqrt{ \frac{1}{n - 1} \sum (x_i - \bar{x})^2 }
$$

Where:

-   $s$ : sample standard deviation\
-   $x_i$ : observed value for the $i$th observation\
-   $\bar{x}$ : sample mean\
-   $(x_i-\bar{x})^2$ : squared deviation from the mean\
-   $n-1$ : denominator used for the sample-based estimate of variability
:::

------------------------------------------------------------------------

### Standard Deviation and Interquartile Range

::: {style="font-size:0.67em"}
**Interquartile Range**:

-   width of the central interval containing the middle 50% of ordered observations
-   measure of dispersion
-   summary statistic for skewed variables and data with outliers

$$
IQR = Q_3 - Q_1
$$

Where:

-   $Q_1$ : 25th percentile \[value below which 25% of observations fall\]
-   $Q_3$ : 75th percentile \[value below which 75% of observations fall\]
:::

------------------------------------------------------------------------

### Numerical Summary for IL-6

::: {style="font-size:0.7em"}
```{r, echo=FALSE}
il6_summary <- biomarker_df |>
  summarise(
    mean_il6 = mean(il6),
    median_il6 = median(il6),
    sd_il6 = sd(il6),
    iqr_il6 = IQR(il6),
    q1_il6 = quantile(il6, 0.25),
    q3_il6 = quantile(il6, 0.75),
    min_il6 = min(il6),
    max_il6 = max(il6)
  )

il6_summary
```

**Interpretation**

-   `mean_il6` \[14.7\]: average of IL-6 concentrations across all patients

-   `median_il6` \[11.7\]: 50th percentile of the ordered IL-6 values

    -   point at which half of observations fall below and half above

-   `sd_il6` \[10.6\]: quantifies the overall dispersion of IL-6 values around the mean

    -   variability across the full distribution

-   `iqr_il6` \[12.3\]: range spanned by the central 50% of IL-6 values

    -   measure of variability that reflects the spread of the majority of observations

-   `q1_il6` \[7.26\] and `q3_il6` \[19.5\]: 50% of IL-6 values fall between 7.26 and 19.5 pg/mL

    -   interval captures the central concentration range of the sample

-   `min_il6` \[1.77\] and `max_il6` \[71.7\]: lower and upper bounds of IL-6 in the sample

Example: focus on evaluating IL-6
:::

------------------------------------------------------------------------

### IL-6 Distribution: Histogram

::: {style="font-size:0.67em"}
```{r, echo=FALSE}
plot <- ggplot(biomarker_df, aes(x = il6)) +
  geom_histogram(bins = 24) +
  labs(
    title = "Histogram of IL-6 Concentration",
    x = "IL-6 (pg/mL)",
    y = "Number of patients"
  ) + theme_bw()

plot
```

**Interpretation**

-   right-skewed histogram shows how a continuous biomarker (IL-6 concentration) is distributed across patients
-   tallest bars indicate the IL-6 concentration range where most patients are observed
-   bars at high values represent patients with elevated inflammatory response
-   horizontal spread of the bars shows that IL-6 values range widely
-   long tail extending to the right indicates right-skewness $\rightarrow$ small number of patients have substantially higher IL-6 levels
:::

------------------------------------------------------------------------

### IL-6 Distribution : Density Plot

::: {style="font-size:0.64em"}
```{r, echo = FALSE}
plot <- ggplot(biomarker_df, aes(x = il6)) +
  geom_density(linewidth = 1) +
  labs(
    title = "Density Plot of IL-6 Concentration",
    x = "IL-6 (pg/mL)",
    y = "Density"
  ) + theme_bw()

plot
```

**Interpretation**

-   density plot : smooth representation of how IL-6 values are distributed across patient observations
-   distribution is right-skewed, with most values concentrated around 10–15 pg/mL \[region representing the most common range of observed values\]
-   curve is right-skewed gradually declines as IL-6 increases, showing that higher concentrations occur less frequently
-   values span a wide range, extending from low concentrations to substantially elevated values \[above \~50 pg/mL\]
:::

------------------------------------------------------------------------

### IL-6 Distribution

::: {style="font-size:0.64em"}
```{r, echo=FALSE}
plot <- ggplot(biomarker_df, aes(x = il6)) +
  geom_histogram(aes(y = after_stat(density)), bins = 22) +
  geom_density(linewidth = 1) +
  labs(
    title = "Histogram with Density Overlay for IL-6",
    x = "IL-6 (pg/mL)",
    y = "Density"
  ) + theme_bw()

plot
```

**Interpretation**

-   bars show how IL-6 observations accumulate across intervals of the measurement scale
-   density curve traces the smooth distributional shape implied by those observations
-   agreement between the bars and the curve confirms the main concentration of the sample

Note:

-   `histogram` displays how many observations fall within intervals
-   `density plot` estimates how probability is distributed across a continuous scale
:::

------------------------------------------------------------------------

### IL-6: Boxplot

::: {style="font-size:0.64em"}
```{r}
ggplot(biomarker_df, aes(y = il6)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of IL-6 Concentration",
    y = "IL-6 (pg/mL)",
    x = NULL
  )
```

**Interpretation**

-   boxplot summarizes the distribution of IL-6 concentrations across patient observations using quartiles and the median
-   line inside the box marks the median
-   lower and upper box boundaries mark 𝑄1 and 𝑄3
-   height of the box corresponds to the interquartile range
-   points beyond the whiskers indicate outliers
-   median IL-6 concentration is approximately 11.7 pg/mL, indicating the central location of the distribution
-   interquartile range (IQR ≈ 12.3 pg/mL) : moderate spread among the central observations
-   middle 50% of values lies between $Q_1 \approx 7.26$ pg/mL and $Q_3 \approx 19.5$ pg/mL, defining the primary concentration range
:::

------------------------------------------------------------------------

### IL-6: Histogram to Probability

::: {style="font-size:0.67em"}
-   continuous biomarker such as IL-6 can be treated as a random variable:

$$X=\text{IL-6 concentration [pg/mL]}$$

probability helps answer:

-   how likely are different IL-6 values?
-   where is probability concentrated?
-   how does probability change across the measurement scale?
:::

------------------------------------------------------------------------

### IL-6: Histogram to Probability

::: {style="font-size:0.67em"}
-   histogram approximates how probability is distributed across intervals of IL-6

    -   bar height reflects relative frequency, which approximates probability
    -   dividing counts by total sample size converts counts into approximate probabilities

**Interpretation**

-   regions with taller bars correspond to higher probability density
-   low bars in the right tail indicate low-probability
-   total area across all bars corresponds to probability ≈ 1

$$\text{∑relative frequencies≈1}$$
:::

------------------------------------------------------------------------

### IL-6: Density as a Probability Model

::: {style="font-size:0.67em"}
-   density curve represents an estimate of the probability density function \[PDF\] of IL-6

$$\text{f(x)=density at IL-6 value x}$$

Key properties:

-   $f(x)≥0$
-   total area under the curve equals 1

$$
\int_{-\infty}^{\infty} f(x)\,dx = 1
$$ Interpretation:

-   height of the curve reflects relative likelihood of IL-6 values near x
-   area under the curve over an interval gives probability

Example:

$$
P(10 \le X \le 20) = \int_{10}^{20} f(x)\,dx
$$ - value corresponds to the region where most IL-6 values are concentrated or the probability that X falls in the 10 to 20 interval
:::

------------------------------------------------------------------------

### IL-6: Boxplot and Probability

::: {style="font-size:0.67em"}
-   boxplot summarizes probability using quantiles

Quantiles partition probability:

-   $Q_1$ : 25th percentile → 25% of observations fall below this value
-   median: 50th percentile
-   $Q_3$ : 75th percentile

$$
P(X \le Q_1) = 0.25, \quad P(X \le \text{Median}) = 0.5
$$

Interpretation for IL-6:

-   middle 50% of probability lies between:

$$7.26≤X≤19.5$$ - interval defines the highest-density region of the distribution

Outliers:

-   represent observations in low-probability regions
-   often correspond to clinically extreme cases
:::

------------------------------------------------------------------------

### Summary

::: {style="font-size:0.67em"}
descriptive features define how the sample should be understood before later modeling decisions are made

plots and probabilities describing continuous variables help answer:

-   `mean and median`: describe where the sample is centered\
-   `standard deviation and IQR`: describe how widely values vary\
-   `histograms and density plots`: describe shape and concentration across the scale\
-   `boxplots`: describe ordered spread and tail extension\
-   `interval probabilities`: describe the fraction of values occupying clinically meaningful ranges

metrics help determine whether values are:

-   approximately symmetric\
-   right-skewed or left-skewed\
-   tightly concentrated or broadly dispersed\
-   characterized by extended tails or unusually large values
:::
